---
title: "Homework 1"
subtitle: "PSTAT 131"
author: "John Ferrone"
date: "10/2/2022"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidymodels)
library(corrplot)
library(janitor)

knitr::opts_chunk$set(echo = TRUE)
```

### Machine Learning Main Ideaas

#### Question 1:
*Define supervised and unsupervised learning. What are the difference(s) between them?*

**Supervised learning**: involves a statistical model built for predicting an oautput based on predictor(s). 

**Unsupervised learning**: involves an inputs without the help of a supervising output, meaning the relationships and structure come from the data.

The difference is the existence of an output, or a training set, which can be used to determine the accuracy of the models.

#### Question 2:
*Explain the difference between a regression model and a classification model, specifically in the context of machine learning.*

The difference between the two models is that regression models predict a *continuous* or *qualitative* output while *classification* problems predict a *categorical* or *qualitative* output which is non-numeric.

#### Question 3:
*Name two commonly used metrics for regression ML problems. Name two commonly used metrics for classification ML problems.*

**Regression**:

* MSE (Mean Squared Error) using training and test data sets. Lower MSE implied that the model is a good fit, while high MSE indicates that it is not
(ISLR 29).

* Finding a model which has a both low variance and low bias is important and is another metric which can help determine whether a fit is good (ISLR 33).

**Classification**:

* The Bayes classifier, a simple method of classification which uses predictors to pick and assign observations to the most likely class (ISLR 37).

* The K-Nearest Neighbors classifier estimates the conditional distribution of $Y$ given a predictor $X$ and classifies observations based on the highest *estimated* probability (ISLR 39).

#### Question 4:
*As discussed, statistical models can be used for different purposes. These purposes can generally be classified into the following three categories. Provide a brief description of each.*

**Descriptive models**: used to emphasize a trend in the data, such as a line on a scatterplot (Lecture 1 39)

**Inferential models**: aims to predict $Y$ with minimum reducible error (Lecture 1 39)

**Predictive models**: aims to test theories and state the relationship between outcome and predictor(s) (Lecture 1 39)

#### Question 5:
*Predictive models are frequently used in machine learning, and they can usually be described as either mechanistic or empirically-driven. Answer the following questions.*

* *Define mechanistic. Define empirically-driven. How do these model types differ? How are they similar?*

**Mechanistic**: Mechanistic, or parametric models are distinct due to assumptions made for $f$, such as $\beta_0 + \beta_1 + \dots$ but won't match true unknown $f$. Adding parameters increases flexibility, but too many leads to overfitting (Lecture 1 38).

**Empirically-Driven**: Non-parametric models make no assumptions about $f$, requires a larger number of observations, and are much more flexible by default but again often overfit the data (Lecture 1 38).

* *In general, is a mechanistic or empirically-driven model easier to understand? Explain your choice.*

In general, it is easier to understand a mechanistic model as they tend to be the simpler model types.

* *Describe how the bias-variance tradeoff is related to the use of mechanistic or empirically-driven models.*

Bias-variance trade-off is related to the use of these models as it is an excellent way to determine if a model is truly fit well, or if it is overfitted. An overfitted model would have much larger variance and bias.

#### Question 6:
*A political candidate's campaign has collected some detailed voter history data from their constituents. The campaign is interested in two questions:*

* *Given a voter's profile/data, how likely is it that they will vote in favor of the candidate?*

* *How would a voter's likelihood of support for the candidate change if they had personal contact with the candidate?*

*Classify each question as either predictive or inferential. Explain your reasoning for each.*

1. Predictive, as this is using known data in order to choose the most likely outcome of how they will vote.

2. Inferential, as this is using information known about the candidate which can not necessarily be quantified.

#### Exercise 1:
```{r}
ggplot(mpg, aes(x = hwy)) + 
  geom_histogram()
```

There seems to be two primary clusters of highway miles per gallon, between about 16-20 and 24-28. Outside of these clusters there tends to be much smaller counts of vehicles.

#### Exercise 2:
```{r}
ggplot(mpg, aes(x = hwy, y = cty)) +
  geom_point()
```
There is a clear linear relationship between highway miles per gallon and city miles per gallon, with both increasing as the other does. This means they have a positive relationship and could likely be modeled using a simple linear model.

#### Exercise 3:
```{r}
ggplot(mpg, aes(y = fct_infreq(manufacturer))) + geom_bar()
```
The manufacturer which produced the most cars is Dodge and the least is Lincoln.

#### Exercise 4:
```{r}
ggplot(mpg, aes(x = hwy, group = cyl)) + geom_boxplot()
```
There is a negative trend between highway miles per gallon and cylinders, with `hwy` decreasing as `cyl` increases.

#### Exercise 5
```{r}
numcars = select_if(mpg, is.numeric)
M = cor(numcars)
corrplot(M, method = 'number', type = 'lower')
```
This plot displays various correlations, such as positive ones between `cyl` and `displ` as well as `hwy` and `cty`. Negative correlations exist between `hwy` and `cyl`, `hwy` and `displ`, `cty` and `displ`, as well as `cty` and `cyl`.
