---
title: "Data Memo"
author: "John Ferrone"
date: "2022-09-27"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)

knitr::opts_chunk$set(echo = TRUE)
```

```{r results='hide'}
rawdata <- read_csv('data/nycschooldata.csv')
```

```{r}
head(rawdata)
```

New York City has over a thousand public schools, most of which report statistics on their schools. This data is collected from grades K-12, and is made up of both numeric and categorical variables. For the purpose of this model, only the numeric variables will be considered. The data was taken from a data set published to Kaggle, the link will be at the end of this memo. Before any tidying is done on the data, there are 161 predictors over 1091 observations, but the number of predictors will likely be reduced to a number closer to twenty. Missing data may also reduce the number of observations represented.

The purpose of this model will be to predict the proficiency of each school in the given categories, based on predictors such as `Rigorous Instruction %`, `Percent of Students Chronically Absent`, and `School Income Estimate`. The specific columns being predicted are `Average ELA Proficiency` and `Average Math Proficiency`. A regression model will be used to do this and it will likely be a combination of descriptive and predictive modelling.

I plan on beginning the exploratory analysis this week, and will test models on the data as they are taught to us. It has already been loaded.

* Weeks 1-3: Exploratory analysis
* Weeks 4-6: Testing models
* Weeks 7-10: Assembling the report and testing final models